---
title: "Wisconsin K12 School Data"
author: "Spencer Schien"
date: "`r Sys.Date()`"
output: github_document
always_allow_html: true
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)

library(wisconsink12)
```

# Introduction

The `wisconsink12` package was developed to increase the accessibility of publicly available data regarding K12 schools in Wisconsin and to facilitate analysis of that data.

The data is organized in a relational database structure, where each table has a unique school identifier that facilitates joins.  As of this writing, the following data tables are included:

* `schools` -- This is a list of all schools in Wisconsin that serves as the unique identifier table for the relational database.
* `enrollment` -- This table contains enrollment data for all Wisconsin schools. Disaggregated enrollment data is not available for private schools.
* `forward_exam` -- This table contains forward exam results for public and Choice-participating private schools.  Disaggregated test results are not available for private schools.
* `graduation` -- This table contains graduation and completion data for all public schools.  Private schools are not included here at all because cohort counts are not available for private schools, and so rates cannot be calculated.
* `report_cards` -- This table contains Report Card data from all schools in Wisconsin.

# Installation

As of this writing, there are no plans to submit the `wisconsink12` package to CRAN, so it must be downloaded from City Forward Collective's [GitHub repository](https://github.com/cityforwardcollective).  This can be achieved with the following code.

```{r eval=FALSE}
# The devtools package is required
# to download from GitHub.
# install.packages("devtools")

devtools::install_github("cityforwardcollective/wisconsink12")
```

# Accessing the Data

## Milwaukee School Report Cards

The `wisconsink12` package is built around the school data it makes available.  This data is organized into tables (listed above), and each table contains a school ID field called the `dpi_true_id`.  This field is a concatenation of a school's District Code and its School Code, with an underscore '_' in between.  Both codes are padded with zeros on the left to a length of four, and choice schools are given a District Code of '0000'.  

Once the `wisconsink12` package is loaded, a message will be displayed listing the tables that are available.

```{r}
# Trying to access tables before loading
# the package will result in an error.

# Load the package, then access the tables.

library(wisconsink12)
```

As the output shows, we indeed have access to the five tables listed above.  We can inspect these tables as we would any dataframe.

```{r}
# Inspect the `schools` table

str(schools)
```

If you get further along after loading the package and you forget what tables are available, you can access that information with the `list_tables()` function.

```{r}
# List the tables available in the 
# `wisconsink12` package.

list_tables()
```

# Most Common Uses

There are a few slices of this data that will be a starting point for any analysis the vast majority of the time.  For example, you might want to perform an analysis on Report Card outcomes for Milwaukee schools.  To easily filter for these schools, you can use the `make_mke_rc()` function.

```{r}
# Call `make_mke_rc()` to create two tables:
# `mke_schools` table and `mke_rc` table 

make_mke_rc()

# List environment objects with `ls()`

ls()
```

The primary result of this function is to create `mke_rc`, a table of Report Card data filtered for Milwaukee schools.  A secondary result is the creation of the `mke_schools` table, which contains a unique list of Wisconsin schools along with certain information on those schools such as agency type, last year the school was open, etc.

An important item to note is that private schools participating in a parental choice program in Wisconsin are required to have a report card for their students attending with voucher funds, and they have the option to receive a report card for all of their students.  The functions in the `wisconsink12` package by default filter for the choice students (designated as the *Private - Choice Students* report card in the `report_card_type` field of the `report_cards` table).  This filter can be changed with the `private_type` argument -- setting it as "all" would filter for the *Private - All Students* report card.


```{r}
# make_mke_rc() was run above, 
# with default private_type = "choice"

# Examine unique report card types

unique(mke_rc$report_card_type)

# Call make_mke_rc() again, 
# with default private_type = "s"

make_mke_rc(private_type = "all")

# Examine unique report card types

unique(mke_rc$report_card_type)
```

We see from the two `unique()` calls that `NA` values remain in both instances -- this is because this field didn't exist for the 2015-16 Report Cards since no private schools had scores computed that year.  We also see that the second `unique()` call still shows *Private - Choice Students* among the unique values, even though we set `private_type` to "all".  This is because designating "all" chooses the *Private - All Students* report card type *when it is available* -- as mentioned above, schools only have that report card type if they elect to do so.

```{r message=FALSE, warning=FALSE}
make_mke_rc()

library(knitr) # `knitr` needed for `kable()` below
library(tidyverse) # tidyverse will be used throughout

table(mke_rc$report_card_type, mke_rc$school_year) %>%
  kable()
```

## How many students are in schools that *Meet Expectations*? 

```{r}
# Designate which ratings fall below 'Meeting Expectations'

low_performers <- c("Alternate Rating - Needs Improvement",
                    "Fails to Meet Expectations",
                    "Fails to Meed Expectations^",
                    "Meets Few Expectations",
                    "Meets Few Expectations^")

not_rated <- c("Not Rated",
               "NR-DATA")

meets_expecs <- report_cards %>%
  filter(school_year == "2018-19") %>%
  mutate(quality = ifelse(overall_rating %in% low_performers, "Not Meeting Expectations", 
                          ifelse(overall_rating %in% not_rated, "Not Rated", "Meeting Expectations"))) %>%
  modify_at("quality", factor, levels = c("Not Meeting Expectations",
                                          "Meeting Expectations")) %>%
  group_by(quality) %>%
  summarise(total_enrollment = sum(school_enrollment))

meets_expecs %>%
  ggplot(aes(quality, total_enrollment)) +
  geom_col()
```

***To Be Updated...***
